Urbanization affects web abundance and aggregation of a funnel-weaving spider, *Agelenopsis pennsylvanica*

### 2023-01-09 manuscript/manuscript.Rmd
Final edits before submitting

The two urban habitats that we studied here -- an urban forest and an urban center -- significantly differed by the selected environmental conditions. The urban forest sites had higher tree cover, plant diversity, and patch area. On the other hand, the urban center exhibited higher impervious cover, light at night, surface temperature, road and highway disturbance, and total road length. These urban habitats each host *Agelenopsis pennsylvanica*, but the spatial patterns displayed by this funnel-weaving spider vary considerably. The spiders are more abundant, aggregated, and build lower webs in the urban center. We hypothesized that many of the environmental conditions that are distinct between the habitats may also be influencing such differences. We found that spatial patterns were often retlated to plant species richness and disturbance from road and highway traffic.

@@

Both webs and spiders of *A. pennsylvanica* were more abundant and aggregated in the urban center than the urban forest. Shorter search distances to the focal web corroborates this finding, as previous studies have used search distance as an estimate of abundance [@buckland_2005, @kral_2018].

Higher abundance of animals in urban habitats may reflect a higher tolerance of urban conditions and disturbances. Tolerance of urban ecosystems is dependent on the ecology and behavior of the organism, such as being a generalist consumer [@sol_2013; @bang_2011], and the scale of the measurements being collected [@lowe_2018; @philpott_2014], resulting in conflicting findings in the literature. We have observed *A pennsylvanica* attacking a wide variety of arthropod prey with no evidence of specialization of a particular type (personal observations). Thus, we suggest that *A. pennsylvanica* is a generalist predator. While arthropods are generally found to decline in urban environments [@wagner_2020; @piano_2020], generalists are more likely to be urban exploiters than specialists [@callaghan_2021]. For example, *Nephila plumipes*, a generalist predator, was found to abundantly occupy urban habitats where prey abundance supported greater numbers of spiders [@lowe_2016].

Prey abundance has been repeatedly found to be a strong predictor of spider abundance [@trubl_2012, @parkinson_2020, @harwood_2001]. Although we might hypothesize higher prey availability in the urban forest, we observed far greater potential for competition from other spider species in the urban forest than the urban center. In the urban center, *A. pennsylvanica* appear to dominate spider communities. Prey abundance, then, may play a role in our observed differences in spider abundance between the urban center and the urban forest.

Multiple reports have found spider aggregations in urban habitats [@trubl_2012; @mammola_et_al_2018], another spatial measure that is often related directly or indirectly to the abundance of prey. Territoriality, especially over prey resources, often drives regular (and non-random) spacing of solitary species [@riechert_1974b], but aggregations can form when abundant resources allow for increased tolerance of neighbors [@gillespie_1987]. Light at night in cities attracts insects creating pockets of increased prey abundance for spiders that may lead to spiders aggregating under such light sources [@davies_et_al_2012]. Although we did not measure microhabitat levels of light at night or prey abundance, *A. pennsylvanica* were more aggregated in the urban center where light at night radiance was higher than the urban forest. Thus, future research should test the impact of prey abundance and competition on spider abundance and aggregation in urban habitats.

We also found that the webs of urban center spiders were built lower than webs in the urban forest, and this result was independent of the height of the substrate. Webs may vary in height based on the type of prey being ensnared. We have observed *Agelenopsis* spiders capture flying, hopping, and crawling insects. Arthropod community structure can be highly variable between urban habitats [@bang_faeth_2011], and urban centers and urban forests may support different types of prey items for *A. pennsylvanica*. Lower webs in the urban center may be ideal for more crawling prey while higher webs in the urban forest may better trap flying insects. Alternatively, webs in different habitats may experience different levels of disturbance. In the urban center where lawns are mowed frequently and ornamental plants are regularly pruned, spiders could maintain lower webs to conserve energy from frequent rebuilding.

Additionally, urban forests likely support (and act as a refuge for) many ground dwelling vertebrates, whose movement along the forest floor could frequently damage lower webs. Greater tree coverage in the urban forest predicted higher webs, which may further suggest that small mammal movement could destroy low webs as areas with greater tree cover may provide protection and food resources for small mammals [@sarkar_2022]. Future research should assess prey type available and small animal abundance and movement between these two habitats. We speculate that vibratory noise may promote lower web heights based on a trend that showed lower web heights overall at sites with higher road disturbance. Perhaps vibratory noise is dampened on a plant nearer to the ground, but this idea requires further testing.

@@

We found that webs and spiders were more abundant (in the urban center) and more aggregated as plant species richness decreased. The urban center sites were covered in grass lawns and managed landscaping that usually had few plant species present, but in high abundance. Sites with lower plant diversity may have few plant species that are highly preferred, leading to higher spider abundance. Research has shown mixed but often positive effects of plant diversity on spider abundance [@malumbres_et_al_2013; @koricheva_et_al_2000], suggesting that plant community composition may play a vital, yet context-specific role. To better understand the role that plant species play in abundance would require comparing plant communities between patches used by spiders versus those devoid of spiders. We speculate that greater aggregation on more diverse plant communities may suggest that *A. pennsylvanica* prefers a certain plant or type of plant, especially in the urban center where we observed similar plants clumped in space.

@@

Disturbance from roads and highways also seemed to play a role in spider spatial patterns. We show that webs and spiders are more abundant in urban center sites that had lower highway disturbance. Living in close proximity to high-traffic roads is likely to affect the vibratory environment that the spider is exposed to. *Agelenopsis pennsylvanica* rely heavily on vibrations to capture prey [@riechert_maupin_1998], court potential mates [@singer_2000], and potentially evade predators [@riechert_hedrick_1990]. Thus, we suspect that vibratory noise may impact web site selection, choosing sites that mitigate these effects. Also, selective spiders in urban habitats may choose plants as web substrates that reduce vibratory noise since different plants transmit vibrations differently [@cocroft_et_al_2006]. Although most research on the effects of noise on arthropod behavior has focused on acoustic rather than vibratory noise [@roberts_howard_2022], some studies have found strong negative effects of acoustic and vibratory noise on arthropod behavior [@carosi_et_al_2019; @bunkley_et_al_2017]. A recent study shows that the orb-weaving spider, *Larinioides sclopetarius* is able to detect airborne sound that induces vibrations in the web [@zhou_et_al_2022].

Lower road and highway disturbance was associated with larger spider aggregations in the urban forest, but had no effect on urban center spiders. Perhaps noise negatively affects urban forest spiders -- encouraging spiders to inhabit areas further from noise sources and reducing space between neighbors. Many spiders are thought to be plastic in the structure and tuning of their webs [@mortimer_2016; @blamires_2010], and urban exploiters are often fast to adjust to changing conditions [@kralj_2017], so spiders reared in or descended from spiders in the urban center might be better able to tune their webs to decrease vibrations, although this hypothesis requires further testing. It will also be important to differentiate whether these differences in behavior between environments is based on developmental exposure, behavioral plasticity, or genetic differentiation.

More road length in the 100-meter radius resulted in fewer but more aggregated urban center webs and spiders. Total road length could be another indirect measure of vibratory noise and/or represent dispersal barriers. High traffic roads may increase mortality of dispersing spiders via collisions with vehicles, or passing cars may influence wind currents making aerial dispersal difficult [@martin_et_al_2019]. Higher spider abundance in smaller patches may be more evidence for the role of environmental barriers. Furthermore, small patch areas may be a preferred habitat of edge specialists [@martinson_fagan_2014], as we suspect *A. pennsylvanica* may be.

@@

Lastly, spiders in the urban center were more aggregated at sites with higher spectral radiance (a measure of surface temperature). Urban centers tend to be warmer on average compared to surrounding areas due to heat trapping via pavement in a phenomenon called the Urban Heat Island Effect [@johnson_et_al_2020]. Aggregation formation in warmer sites could be the result of preference for or acclimation to warmer areas [@meineke_et_al_2013] or perhaps aggregation occurs at limited thermal refuges [@riechert_1974b]. We have shown that human-caused environmental changes correlate with spider spatial patterns, but there are likely many other forms of human disturbance than those tested here.

@@

Our study shows that a solitary, generalist predator is more abundant and aggregated with lower web height in an urban center compared to an urban forest -- habitats that vary significantly in environmental conditions. We did not measure community-level interactions among organisms, such as the effects of prey abundance. However, correlations between spatial pattern and environmental condition indirectly indicate that such interspecific interactions may be essential for monitoring spider aggregations. Spider aggregations in urban centers can have medical consequences, such as dense *Lactrodectus hesperus* aggregations [@johnson_et_al_2011], and aesthetic concerns, as is the case with aggregations of *Brigittea civica* under archways of historic buildings [@mammola_et_al_2018]. On the other hand, spider aggregations could control pest outbreaks that threaten human health and food supply. We emphasize that elucidating the complex relationship between the arthropod abundance and the impacts of urbanization is essential for our understanding of how urbanization affects ecosystem services and food web interactions. Our variables that estimate the potential for vibratory noise negatively affected spider abundance while promoting lower web heights, potentially to reduce interference from vibratory noise. Noise variables had opposite effects on aggregative behavior of spiders in the different habitats that may indicate differences in tolerance to noise, but more research is necessary to test this hypothesis. Thus, urban habitats can vary in environmental conditions which can determine how animals distribute themselves in space and affect predator-prey interactions.

@@

The authors have no relevant financial or non-financial interests to disclose.

@@

>>>>

The two urban habitats that we studied here -- an urban center and an urban forest -- significantly differed by the selected environmental conditions. The urban center exhibited higher impervious cover, light at night, surface temperature, road and highway disturbance, and total road length. On the other hand, the urban forest sites had higher tree cover, plant diversity, and patch area. These urban habitats each host *Agelenopsis pennsylvanica*, but the spatial patterns displayed by this funnel-weaving spider vary considerably. The spiders are more abundant, aggregated, and build lower webs in the urban center. We hypothesized that many of the environmental conditions that are distinct between the habitats may also be influencing such differences. We found that spatial patterns were often related to plant species richness and disturbance from road and highway traffic.

@@

Both webs and spiders of *A. pennsylvanica* were more abundant and aggregated in the urban center than the urban forest. Shorter search distances to the focal web corroborate this finding, as previous studies have used search distance as an estimate of abundance [@buckland_2005, @kral_2018].

Higher abundance of animals in urban habitats may reflect a higher tolerance of urban conditions and disturbances. Tolerance of urban ecosystems is dependent on the ecology and behavior of the organism, such as being a generalist consumer [@sol_2013; @bang_2011], and the scale of the measurements being collected [@lowe_2018; @philpott_2014]. We have observed *A pennsylvanica* attacking a wide variety of arthropod prey with no evidence of specialization of a particular type (personal observations). Thus, we suggest that *A. pennsylvanica* is a generalist predator. While arthropods are generally found to decline in urban environments [@wagner_2020; @piano_2020], generalists are more likely to be urban exploiters than specialists [@callaghan_2021]. For example, *Nephila plumipes*, a generalist predator, was found to abundantly occupy urban habitats where prey abundance supported greater numbers of spiders [@lowe_2016]. Prey abundance has been repeatedly found to be a strong predictor of spider abundance [@harwood_2001; @parkinson_2020; @trubl_2012]. Although we might hypothesize higher prey availability in the urban forest, we observed far greater potential for competition from other spider species in the urban forest than the urban center. In the urban center, *A. pennsylvanica* appear to dominate spider communities. Prey abundance, then, may play a role in our observed differences in spider abundance between the urban center and the urban forest.

Multiple reports have found spider aggregations in urban habitats [@trubl_2012; @mammola_et_al_2018], another spatial measure that is often related directly or indirectly to the abundance of prey. Territoriality, especially over prey resources, often drives regular (and non-random) spacing of solitary species [@riechert_1974b], but aggregations can form when abundant resources allow for increased tolerance of neighbors [@gillespie_1987]. Light at night in cities attracts insects, creating pockets of increased prey abundance for spiders that may lead to spiders aggregating under such light sources [@davies_et_al_2012]. Although we did not measure microhabitat levels of light at night or prey abundance, *A. pennsylvanica* were more aggregated in the urban center where light at night radiance was higher than the urban forest. Thus, future research should test the impact of prey abundance and competition on spider abundance and aggregation in urban habitats.

We also found that the webs of urban center spiders were built lower than webs in the urban forest, and this result was independent of the height of the substrate. One possible explanation for this pattern is that vibratory noise promotes lower web heights. Vibratory noise might, for example, be dampened on a plant nearer to the ground. We did find a trend that showed lower web heights overall at sites with higher road disturbance, but this idea requires further testing. Another potential explanation is that webs may vary in height based on the type of prey being ensnared. We have observed *Agelenopsis* spiders capture flying, hopping, and crawling insects. Arthropod community structure can be highly variable between urban habitats [@bang_faeth_2011], and urban centers and urban forests may support different types of prey items for *A. pennsylvanica*. Lower webs in the urban center may be ideal for more crawling prey while higher webs in the urban forest may intercept more flying insects. Alternatively, webs in different habitats may experience different levels of disturbance. In the urban center where lawns are mowed frequently and ornamental plants are regularly pruned, spiders could maintain lower webs to conserve energy from frequent rebuilding. Additionally, urban forests likely support (and act as a refuge for) many ground dwelling vertebrates, whose movement along the forest floor could frequently damage lower webs. Greater tree coverage in the urban forest predicted higher webs, which may further suggest that small mammal movement could destroy low webs as areas with greater tree cover may provide protection and food resources for small mammals [@sarkar_2022]. Future research should assess the prey types available and small animal abundance and movement between these two habitats.

@@

We found that webs and spiders were more abundant (in the urban center) and more aggregated as plant species richness decreased. The urban center sites were covered in grass lawns and managed landscaping that usually had few plant species present, but in high abundance. Sites with lower plant diversity may have few plant species that are highly preferred, leading to higher spider abundance. Research has shown mixed but often positive effects of plant diversity on spider abundance [@malumbres_et_al_2013; @koricheva_et_al_2000], suggesting that plant community composition may play a vital, yet context-specific role in spider spatial patterns. To better understand the role that plant species play in abundance would require comparing plant communities between patches used by spiders versus those devoid of spiders. We speculate that greater aggregation on more diverse plant communities may suggest that *A. pennsylvanica* prefers a certain plant or type of plant, especially in the urban center where we observed similar plants clumped in space.

@@

Disturbance from roads and highways also seemed to play a role in spider spatial patterns. We show that webs and spiders are more abundant in urban center sites that had lower highway disturbance. Living in close proximity to high-traffic roads is likely to affect the vibratory environment that the spider is exposed to. *Agelenopsis pennsylvanica* rely heavily on vibrations to capture prey [@riechert_maupin_1998], court potential mates [@singer_2000], and potentially evade predators [@riechert_hedrick_1990]. Thus, we suspect that vibratory noise may impact web site selection, with spiders choosing sites that mitigate these effects. Also, selective spiders in urban habitats may choose plants as web substrates that reduce vibratory noise since various plants transmit vibrations differently [@cocroft_et_al_2006]. Although most research on the effects of noise on arthropod behavior has focused on acoustic rather than vibratory noise [@roberts_howard_2022], some studies have found strong negative effects of acoustic and vibratory noise on arthropod behavior [@carosi_et_al_2019; @bunkley_et_al_2017]. A recent study shows that the orb-weaving spider, *Larinioides sclopetarius* is able to detect airborne sound that induces vibrations in the web [@zhou_et_al_2022].

Lower road and highway disturbance was associated with larger spider aggregations in the urban forest, but had no effect on urban center spiders. Perhaps noise negatively affects urban forest spiders -- encouraging spiders to inhabit areas further from noise sources and reducing space between neighbors. Many spiders are thought to be plastic in the structure and tuning of their webs [@mortimer_2016; @blamires_2010], and urban exploiters are often fast to adjust to changing conditions [@kralj_2017]. As such, spiders reared in or descended from spiders in the urban center might be better able to tune their webs to decrease vibrations, although this hypothesis requires further testing. It will also be important to differentiate whether these differences in behavior between environments are based on developmental exposure, behavioral plasticity, or genetic differentiation.

More road length in the 100-meter radius correlated with fewer but more aggregated urban center webs and spiders. Total road length could be another indirect measure of vibratory noise and/or represent dispersal barriers. High-traffic roads may increase mortality of dispersing spiders via collisions with vehicles, or passing cars may influence wind currents making aerial dispersal difficult [@martin_et_al_2019]. Higher spider abundance in smaller patches may be more evidence for the role of environmental barriers. Furthermore, small patch areas may be a preferred habitat of edge specialists [@martinson_fagan_2014], as we suspect *A. pennsylvanica* may be.

@@

Lastly, spiders in the urban center were more aggregated at sites with higher spectral radiance (a measure of surface temperature). Urban centers tend to be warmer on average compared to surrounding areas due to heat-trapping via pavement in a phenomenon called the Urban Heat Island Effect [@johnson_et_al_2020]. Aggregation formation in warmer sites could be the result of preference for or acclimation to warmer areas [@meineke_et_al_2013] or perhaps aggregation occurs at limited thermal refuges [@riechert_1974b].

@@

We have shown that human-caused environmental changes correlate with spider spatial patterns, but there are likely many other forms of human disturbance than those tested here. Our study shows that a solitary, generalist predator (*A. pennsylvanica*) is more abundant and aggregated with lower web height in an urban center compared to an urban forest -- habitats that vary significantly in environmental conditions. We did not measure community-level interactions among organisms, such as the effects of prey abundance; however, correlations between spatial pattern and environmental conditions indirectly indicate that such interspecific interactions may be important influencers of spider aggregations. Noise may also greatly influence spider spatial patterns. Our variables that estimate the potential for vibratory noise negatively affected spider abundance while promoting lower web heights, potentially to reduce interference from vibratory noise. Noise variables also had opposite effects on aggregative behavior of spiders in the different habitats. These differences may indicate variability in tolerance to noise, but more research is necessary to test this hypothesis.

Spider aggregations in urban centers are important to understand, as they can have medical consequences -- e.g., dense black widow (*Lactrodectus hesperus*) aggregations [@johnson_et_al_2011] -- and aesthetic concerns -- e.g., aggregations of the civic meshweaver (*Brigittea civica*) under archways of historic buildings [@mammola_et_al_2018]. On the other hand, spider aggregations could control pest outbreaks that threaten human health and food supply. We emphasize that elucidating the complex relationship between arthropod abundance and the impacts of urbanization is essential for our understanding of how urbanization affects ecosystem services and food web interactions.

@@

The authors have no competing interests to declare that are relevant to the content of this article.

@@

All authors contributed to the study conception and design. Data collection was performed by Brandi Pessman, Madison Hays, and Earl Agpawa under the supervision of Eileen Hebets. Formal analysis and the first draft of the manuscript were completed by Brandi Pessman. All authors reviewed and edited subsequent versions of the manuscript and approved its final draft.
